Fused regression for network inference in multiple species

### 2015-08-31 paper1.tex

\documentclass[11pt]{article}
\usepackage{graphicx}
%\usepackage{mathtools}
\usepackage{amssymb,amsmath}
\usepackage{textcomp}
\DeclareMathOperator*{\argmin}{arg\,min}

@@

\section{Introduction}
As the volume and variety of genome scale data have exploded, the goal of accurately modeling gene regulatory networks has become more attainable. Large-scale data collection efforts have contributed to the development of high quality networks, but there remain many biological processes and organisms for which there does not exist enough data. Furthermore, as new technologies are developed and some old ones are replaced, such as RNAseq and microarray, it becomes important to be able to combine different data sets, lest we lose valuable information from previous studies. The problems of inferring networks in related species, processes, and using different technologies likely are not independent ones, and benefit from the infusion of additional data.

The extent to which datasets are related is an important consideration when deciding to combine them. Abundant data in one species should be a useful source when learning the network of a related species. But often, how much and what of orthologous networks' structure is conserved, is not known a priori. We would like, then to be able to learn the conserved subnetwork in order to make better predictions using multi-source network inference.

We present two methods for network inference which use fused regression, a type of penalized least squares regression, in which the difference of regulatory weights which are thought to be similar are penalized. Our fused L2 method uses a ridge fusion penalty on these differences, and is useful where the relationships between nodes (similarity of genes between data sets) is reliable. Our adaptive fusion uses a non-concave saturating fusion penalty to simultaneously learn multiple networks and conservation of these networks.

To learn gene regulatory networks, we employ a linear model of gene expression in which the rate of expression is assumed to be proportional to a weighted sum of transcription factor expression. These weights represent the regulatory interactions we wish to learn. We can fit these weights simultaneously in multiple data sources of interest using time series and steady state expression data. This is similar to many existing network inference methods and produces identical answers to fitting each network separately. To take advantage of cross-dataset information, we introduce additional constraints on the problem that penalize differences between regulatory weights which are expected to be similar. Genes are not always regulated by the same transcription factors under different conditions. For example, a TF which promotes expression of a gene under a given process may repress that same gene at another point. Using data from these conditions to learn one network will prevent the identification of these contradictory regulatory interactions; it is therefore important to be able to combine data in a principled manner. We can use knowledge about the conservation of regulation while leveraging the additional data to learn networks describing disparate processes. In the case of multiple species, there has been a great deal of work showing that functional conservation exists in gene regulatory networks even across large evolutionary distance. We make the assumption that if a transcription factor in one species is orthologous to a TF in a related species, and a gene in the first species is orthologous to a gene in the second, then the interaction between the TF and gene in the first species should be similar to the interaction between the orthologous TF, gene pair in the second species. By utilizing information about conservation of gene function across data set, species, and biological process, we enable the usage of less directly relevant data and improve network inference.

There exist several tools for predicting orthology between genes using sequence similarity. Although sequence similarity is a useful predictor of functional similarity, we recognize that some genes have taken on different functions and therefore may have new regulatory interactions. When this occurs, the benefits in overall performance from our L2 fusion approach come at the cost of impaired prediction of regulators for these genes. The constraints we impose favor weight configurations in which fused interaction weights are similar to one another. Our fused L2 method minimizes a cost function that strives to simultaneously fit expression data in each species and produce networks that are consistent with these evolutionary constraints. When a gene's regulators have changed through evolution, there may be no way of achieving both these goals. Similarly, if a gene has different regulators during different processes, the expression data may conflict with the fusion constraints. In these cases, the network inference procedure would be improved by relaxing or ignoring these evolutionary constraints. However, we don't know a priori which constraints to enforce and which to relax.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[scale=.5,trim=0mm 0mm 0mm 0mm,clip]{direct_solve.eps}
\end{figure}

>>>>

\documentclass[11pt]{article}
\usepackage{graphicx}
\usepackage[pdf]{pstricks}
%\usepackage{mathtools}
\usepackage{amssymb,amsmath}
\usepackage{textcomp}
\DeclareMathOperator*{\argmin}{arg\,min}

@@

\section{Introduction}
As the volume and variety of genome scale data have exploded, the goal of accurately modeling gene regulatory networks has become more attainable. Large-scale data collection efforts have contributed to the development of high quality networks, but there remain many biological processes and organisms for which there does not exist adequate data. Furthermore, as new technologies are developed and some old ones are replaced, such as RNAseq and microarray, it becomes important to be able to combine data from multiple platforms, lest we lose valuable information from previous studies. The problem of inferring related -- but not necessarily identical -- structure from related -- but not identical -- data is ubiquitous in biology. Multi-source network inference has applications for learning networks in multiple related species, for learning networks associated with distinct processes within the same species, and for learning networks based on heterogenous data sources.

Most algorithms for network inference can not use data from multiple sources. This is particularly problematic when studying a lesser used model organism, or a biological process that is not widely studied. Approaches that do utilize data from multiple sources have largely relied on either learning individual networks and combining the solutions, which is unviable when one or more of the data sources is very limited, or on treating the data as if it came from a single source, which is limited to sources with a high degree of overlap (ie species with many genes in common). Moreover, combining the data into a single source assumes that the underlying networks are identical. In reality, the similarity of networks in related species is unknown. A method is needed that can flexibly combine data sources based on their similarity.

We present two methods for network inference based on linear estimates of gene expression dynamics, extending existing linear dynamical-systems methods for network inference. The core of both methods is the observation that biological information about the relatedness of two networks can be used to guess which network coefficients should be similar to one another in a multi-source network inference problem (ie orthologous TFs should regulate orthologous genes), and that these constraints can be efficiently represented as penalties in a least-squares regression problem. Our first method -- fused ridge -- method uses an L2 penalty on the differences between a priori similar interactions, and is useful where the relationships between networks (similarity of genes between data sets) is reliable. In the case where both networks contain an identical set of genes and TFs, this approach can be thought of as parametrically interpolating between treating the data sources separately and combining them together. However, our method allows useful pooling of data even when . Our second method -- adaptive fusion -- uses a non-concave saturating fusion penalty to simultaneously learn the constrained networks and to learn which constraints should be relaxed (ie which parts of the network are genuinely different).

In the case of multiple species, there has been a great deal of work showing that functional conservation exists in gene regulatory networks even across large evolutionary distance. We make the assumption that if a pair of TFs are orthologous to one another, then their regulatory effects on a pair of orthologous genes should be similar. This orthology information forms the basis of a large number of constraints on the network inference problems, which allow data from one organism to improve network inference on another organism. This very general framework for multi-species network inference can be extended to large numbers of distinct organisms, each contributing data with only partially overlapping sets of genes.

Orthology -- the measure of gene similarity we use -- is based on readily identifiable sequence similarity. Although sequence similarity is a useful predictor of functional similarity, we recognize that some genes have taken on different functions and therefore may have new regulatory interactions. These cases of neofunctionalization can be difficult to verify by simply comparing independently inferred networks; because network inference is typically underconstrained, there could be other plausible networks in which the coefficients being compared are similar. This is in contrast with our multi-source method: because the constraints we impose favor weight configurations in which fused interaction weights are similar to one another, a solution in which a priori similar regulatory interactions are found to be different is strong evidence that they really are different. Our fused L2 method minimizes a cost function that strives to simultaneously fit expression data in each species and produce networks that are consistent with these evolutionary constraints. When a gene's regulators have changed through evolution, there may be no way of achieving both these goals. Similarly, if a gene has different regulators during different processes, the expression data may conflict with the fusion constraints. In these cases, the network inference procedure would be improved by relaxing or ignoring these evolutionary constraints. However, we don't know a priori which constraints to enforce and which to relax.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[scale=.5,trim=0mm 0mm 0mm 0mm,clip]{direct_solve.png}
\end{figure}
